Numerical modelling of the Cordilleran ice sheet

### 2014-07-20 kappa.tex
Rework the abstract.

This doctoral thesis presents a study of the late glacial history of
north-western North America using numerical ice sheet modelling calibrated by
geological evidence.

The North American Cordillera is an immense maze of mountain ranges and
intermontane basins which once hosted an ice mass comparable to Greenland: the
Cordilleran ice sheet. Although less extensive than the former Laurentide and
Fennoscandian ice sheets, the Cordilleran ice sheet perhaps remains the least
understood Pleistocene ice sheet in terms of its extent, volume, and dynamics.
Because of the irregular topography of the ice sheet's subglacial landscape,
ice flowed along tortuous pathways, sometimes following the major topographical
troughs, and sometimes flowing across them.

This thesis presents simulations performed with PISM, a numerical ice sheet
model that approximates glacier flow by basal sliding and internal deformation.
These simulations show that the somewhat paradoxical geometry of the last
glacial maximum Cordilleran ice sheet can be largely explained by the present
distribution of temperature and precipitation in the region combined with
paleoclimate reconstructions, provided an
accurate treatment of seasonality, orographic precipitation, and daily
temperature variability in the ice sheet model. Nevertheless, at the scale of
the last glacial cycle, this configuration appears ephemeral. In contrast, the
average stage of glaciation consists of isolated glaciers and ice caps covering
the major mountain ranges of the North American Cordillera.

The simulations presented in this thesis demonstrate the potential of combining
numerical ice sheet modelling with analysis of geological evidence in order to
improve our understanding of former glaciations.

>>>>

This doctoral thesis presents a study of the glacial history of northwestern
North America using numerical ice sheet modelling calibrated against field
evidence. The area is characterized by the steep topography of several mountain
ranges separated by large inter-montane depressions that were once covered by a
large-scale ice mass, the former Cordilleran ice sheet. Due to this complex
topographic setting, the Cordilleran ice sheet remains the least understood
Pleistocene ice sheet in terms of its extent, volume, and dynamics. Because of
the irregular topography on which the ice sheet formed, field studies have
often only had local or regional relevance. Numerical modelling, on the other
hand, allows quantitative reconstructions of whole ice sheet evolution based on
approximated physics of glacier flow.

This thesis presents simulations performed with PISM, a numerical ice sheet
model that approximates glacier flow by basal sliding and internal deformation,
and allows simulations of ice sheet growth and decay over multi-millenial time
scales. Sensitivity experiments have shown that simulations adopting climate
forcing from the North American Regional Reanalysis (NARR) yield ice sheet
extents that best fit the field constraints during the Last Glacial Maximum
(LGM). These simulations show that the geometry of the LGM Cordilleran ice
sheet can be largely explained by present-day temperature and precipitation
fields combined with palaeo-climate reconstructions if an accurate treatment of
seasonality, orographic precipitation, and daily temperature variability is
included in the ice sheet model.

In contrast to this short-lived, maximum ice sheet configuration, most of the
last glacial cycle is characterized by an intermediate state of glaciation
including isolated glaciers and ice caps covering the major mountain ranges of
the North American Cordillera. The largest and most persistent of these ice
sheet nucleation centres is located over the Skeena Mountains. Modelled
deglaciation conditions broadly match field evidence, and where they diverge,
such as over the Fraser Plateau, they indicate either a lack of important
processes in the model, or a more complex glaciation history than previously
thought. The model yields a readvance of the ice sheet during late glacial
times, which appears to match field evidence, before final retreat towards the
Skeena Mountains, which we identify as a key area to understanding glacial
dynamics of the Cordilleran ice sheet, highlighting the need for further
geological investigation of this region.

The simulations presented in this thesis demonstrate the potential of numerical
ice sheet modelling to inform on ice sheet history over glacial cycles,
provided that ice sheet models can be calibrated against field constraints.
